---
title: "Concept induction with lloomr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concept induction with lloomr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lloomr)
set.seed(1)
```

lloomr is an R implementation of the **LLooM** concept-induction algorithm
(Lam et al., 2024, CHI; Python package `text_lloom`). From a collection of
texts, it induces interpretable **concepts** — each a short name plus a
one-sentence inclusion criterion — and then scores every document against
every concept.

The pipeline:

1. **Distill** — extract salient quotes ([distill_filter()], optional) and
   compress to bullet points ([distill_summarize()]).
2. **Cluster** — embed bullets, reduce with UMAP, cluster with HDBSCAN
   ([cluster_texts()]).
3. **Synthesize** — an LLM proposes named concepts per cluster
   ([synthesize_concepts()]).
4. **Review** — remove weak concepts, merge overlaps, select the best
   ([review_concepts()]).
5. **Score** — rate every document against every concept on a 5-point
   scale ([score_concepts()]).
6. **Refine / loop** — drop generic or rare concepts
   ([refine_concepts()]), find uncovered documents ([loop_docs()]).

LLM calls go through [ellmer](https://ellmer.tidyverse.org) chat objects,
so any supported provider works; responses are constrained by structured
output schemas, not parsed out of free text.

## The session pipeline

The chunks below are not evaluated in this vignette (they call LLM APIs);
they show the canonical workflow. With an `OPENAI_API_KEY` set, they run
as-is.

```{r pipeline, eval = FALSE}
library(lloomr)

# df has one row per document, with columns "text" and "doc_id"
sess <- lloom_session(df, text_col = "text", id_col = "doc_id")

# Estimate cost before spending anything
lloom_estimate_gen_cost(sess)

# Distill -> cluster -> synthesize -> review; activate the best 8 concepts.
# `seed` steers generation toward a topic of interest (optional);
# `sample_n` generates concepts from a sample while scoring stays full.
sess <- lloom_gen(sess, seed = "media trust", max_concepts = 8, sample_n = 200)
sess$concepts

# Score every document against the active concepts
sess <- lloom_score(sess)
score_df <- lloom_results(sess)

# Time, tokens, and dollars per step
summary(sess)
```

Every step is also a standalone function on plain data frames
(`distill_summarize()`, `cluster_texts()`, `synthesize_concepts()`, ...) —
the session is a convenience, not a requirement.

## Multi-label scores, single-label topics

Scoring is **multi-label**: each (document, concept) pair is rated
independently, so a document can match several concepts or none. When a
mutually exclusive partition is needed, freeze a topic set and use one of:

```{r single-label, eval = FALSE}
# Forced-choice LLM classification (labels constrained to your topic set)
assignments <- assign_topics(df, "text", "doc_id", sess$concepts, chat)

# Or: free, deterministic argmax over existing scores
slotted <- slot_by_score(score_df, "doc_id", threshold = 1)
```

## Visualizing results

[lloom_vis()] draws the concept-by-group heatmap (the R replacement for
the Python package's interactive matrix widget). It needs only a scored
session, so we demonstrate with a small synthetic one — no API calls:

```{r vis-demo, fig.width = 6.5, fig.height = 3, eval = requireNamespace("ggplot2", quietly = TRUE)}
docs <- data.frame(
  doc_id = as.character(1:8),
  text = paste("document", 1:8),
  party = rep(c("D", "R"), each = 4)
)
concepts <- new_concepts(
  name = c("Media Distrust", "Vaccine Promotion"),
  prompt = c("Distrusts media?", "Promotes vaccines?"),
  active = TRUE
)

# A synthetic score grid (in practice: lloom_score() / score_concepts())
score_df <- expand.grid(
  doc_id = docs$doc_id, concept_name = concepts$name,
  stringsAsFactors = FALSE
)
score_df$concept_id <- concepts$id[match(score_df$concept_name, concepts$name)]
score_df$text <- docs$text[match(score_df$doc_id, docs$doc_id)]
score_df$score <- c(1, 1, 0, 0, 1, 1, 1, 0,   # Media Distrust
                    0, 1, 1, 0, 0, 0, 0, 0)   # Vaccine Promotion
score_df$rationale <- ""
score_df$highlight <- ""
score_df$concept_seed <- NA_character_

sess <- lloom_session(docs, "text", "doc_id",
                      distill_chat = "unused", synth_chat = "unused",
                      score_chat = "unused")
sess$concepts <- concepts
sess$score_df <- score_df

lloom_vis(sess, slice_col = "party")
```

The underlying tidy matrix is available via [concept_matrix()] for custom
plots, and [lloom_export()] produces a per-concept evidence table
(criteria, prevalence, highlight quotes).

## Further reading

- Lam, M., et al. (2024). *Concept Induction: Analyzing Unstructured Text
  with High-Level Concepts Using LLooM.* CHI 2024.
  <https://doi.org/10.1145/3613904.3642830>
- Upstream Python package: <https://github.com/michelle123lam/lloom>
