---
title: "Concept induction with lloomr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concept induction with lloomr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lloomr)
set.seed(1)
```

lloomr is an R implementation of the **LLooM** concept-induction algorithm
(Lam et al., 2024, CHI; Python package `text_lloom`). From a collection of
texts, it induces interpretable **concepts** — each a short name plus a
one-sentence inclusion criterion — and then scores every document against
every concept.

The pipeline:

1. **Distill** — extract salient quotes ([distill_filter()], optional) and
   compress to bullet points ([distill_summarize()]).
2. **Cluster** — embed bullets, reduce with UMAP, cluster with HDBSCAN
   ([cluster_texts()]).
3. **Synthesize** — an LLM proposes named concepts per cluster
   ([synthesize_concepts()]).
4. **Review** — remove weak concepts, merge overlaps, select the best
   ([review_concepts()]).
5. **Score** — rate every document against every concept on a 5-point
   scale ([score_concepts()]).
6. **Refine / loop** — drop generic or rare concepts
   ([refine_concepts()]), find uncovered documents ([loop_docs()]).

LLM calls go through [ellmer](https://ellmer.tidyverse.org) chat objects,
so any supported provider works; responses are constrained by structured
output schemas, not parsed out of free text.

## The session pipeline

The chunks below are not evaluated in this vignette (they call LLM APIs);
they show the canonical workflow. With an `OPENAI_API_KEY` set, they run
as-is.

```{r pipeline, eval = FALSE}
library(lloomr)

# df has one row per document, with columns "text" and "doc_id"
sess <- lloom_session(df, text_col = "text", id_col = "doc_id")

# Estimate cost before spending anything
lloom_estimate_gen_cost(sess)

# Distill -> cluster -> synthesize -> review; activate the best 8 concepts.
# `seed` steers generation toward a topic of interest (optional);
# `sample_n` generates concepts from a sample while scoring stays full.
sess <- lloom_gen(sess, seed = "media trust", max_concepts = 8, sample_n = 200)
sess$concepts

# Score every document against the active concepts
sess <- lloom_score(sess)
score_df <- lloom_results(sess)

# Time, tokens, and dollars per step
summary(sess)
```

Every step is also a standalone function on plain data frames
(`distill_summarize()`, `cluster_texts()`, `synthesize_concepts()`, ...) —
the session is a convenience, not a requirement.

## Choosing your LLM (provider and models)

The model decision is made **when you create the session** (or, for
standalone operators, via their `chat` argument) — `lloom_gen()` and
`lloom_score()` deliberately take no model argument; they use what the
session holds. You choose the provider by choosing the ellmer constructor:

```{r providers, eval = FALSE}
# One model for everything — OpenAI, Anthropic, Gemini, or local:
sess <- lloom_session(df, "text", "doc_id",
  chat = ellmer::chat_openai(model = "gpt-5.2", echo = "none"))

sess <- lloom_session(df, "text", "doc_id",
  chat = ellmer::chat_anthropic(model = "claude-sonnet-4-6", echo = "none"))

sess <- lloom_session(df, "text", "doc_id",
  chat = ellmer::chat_ollama(model = "llama3.3", echo = "none"))  # local

# Or mix per step — cheap for high-volume distill/score, capable for
# synthesis (this mirrors the package defaults):
sess <- lloom_session(df, "text", "doc_id",
  distill_chat = ellmer::chat_openai(model = "gpt-5.4-nano", echo = "none"),
  synth_chat   = ellmer::chat_openai(model = "gpt-5.2", echo = "none"),
  score_chat   = ellmer::chat_openai(model = "gpt-5.4-nano", echo = "none"))
```

If you pass nothing, the defaults are gpt-5.4-nano (distill, score) and
gpt-5.2 (synthesis), which require `OPENAI_API_KEY`.

One caveat: **clustering uses OpenAI embeddings by default** regardless of
your chat provider (some providers, like Anthropic, have no embeddings
API). To go fully non-OpenAI, supply your own embedding function:

```{r embed-fn, eval = FALSE}
sess <- lloom_session(df, "text", "doc_id",
  chat = ellmer::chat_anthropic(model = "claude-sonnet-4-6", echo = "none"),
  embed_fn = function(texts) my_embedding_matrix(texts))  # any fn: texts -> matrix
```

## Bring your own concepts (scoring a human codebook)

Concept *generation* is optional. If you already have a codebook —
theory-driven categories, concepts from a previous study, or hand-edited
output of an earlier run — you can use lloomr purely **deductively**:
write the concepts yourself and score every document against them. Each
concept just needs a name and a yes/no inclusion-criterion question:

```{r byo-concepts, eval = FALSE}
codebook <- new_concepts(
  name = c("Economic Anxiety", "Media Distrust"),
  prompt = c(
    "Does the text express concern about economic conditions?",
    "Does the text express distrust toward news media?"
  ),
  active = TRUE
)

chat <- ellmer::chat_openai(model = "gpt-5.4-nano", echo = "none")
score_df <- score_concepts(df, "text", "doc_id", codebook, chat,
                           get_highlights = TRUE)
```

The same works inside a session — skip `lloom_gen()` entirely and add
concepts by hand:

```{r byo-session, eval = FALSE}
sess <- lloom_session(df, "text", "doc_id")
sess <- lloom_add_concept(sess, "Economic Anxiety",
                          "Does the text express concern about economic conditions?")
sess <- lloom_add_concept(sess, "Media Distrust",
                          "Does the text express distrust toward news media?")
sess <- lloom_score(sess)
```

And for mutually exclusive categories, [assign_topics()] accepts a plain
character vector of topic names — no concept objects needed at all.

## Multi-label scores, single-label topics

Scoring is **multi-label**: each (document, concept) pair is rated
independently, so a document can match several concepts or none. When a
mutually exclusive partition is needed, freeze a topic set and use one of:

```{r single-label, eval = FALSE}
# Forced-choice LLM classification (labels constrained to your topic set)
assignments <- assign_topics(df, "text", "doc_id", sess$concepts, chat)

# Or: free, deterministic argmax over existing scores
slotted <- slot_by_score(score_df, "doc_id", threshold = 1)
```

## Visualizing results

[lloom_vis()] draws the concept-by-group heatmap (the R replacement for
the Python package's interactive matrix widget). It needs only a scored
session, so we demonstrate with a small synthetic one — no API calls:

```{r vis-data}
docs <- data.frame(
  doc_id = as.character(1:8),
  text = paste("document", 1:8),
  party = rep(c("D", "R"), each = 4)
)
concepts <- new_concepts(
  name = c("Media Distrust", "Vaccine Promotion"),
  prompt = c("Distrusts media?", "Promotes vaccines?"),
  active = TRUE
)

# A synthetic score grid (in practice: lloom_score() / score_concepts())
score_df <- expand.grid(
  doc_id = docs$doc_id, concept_name = concepts$name,
  stringsAsFactors = FALSE
)
score_df$concept_id <- concepts$id[match(score_df$concept_name, concepts$name)]
score_df$text <- docs$text[match(score_df$doc_id, docs$doc_id)]
score_df$score <- c(1, 1, 0, 0, 1, 1, 1, 0,   # Media Distrust
                    0, 1, 1, 0, 0, 0, 0, 0)   # Vaccine Promotion
score_df$rationale <- ""
score_df$highlight <- ""
score_df$concept_seed <- NA_character_

sess <- lloom_session(docs, "text", "doc_id",
                      distill_chat = "unused", synth_chat = "unused",
                      score_chat = "unused")
sess$concepts <- concepts
sess$score_df <- score_df
```

```{r vis-demo, fig.width = 6.5, fig.height = 3, eval = requireNamespace("ggplot2", quietly = TRUE)}
lloom_vis(sess, slice_col = "party")
```

The underlying tidy matrix is available via [concept_matrix()] for custom
plots, and [lloom_export()] produces a per-concept evidence table
(criteria, prevalence, highlight quotes).

## Saving your results

The score table is a **plain data frame** — one row per (document,
concept) pair — so saving it is one ordinary line:

```{r save-csv, eval = FALSE}
readr::write_csv(lloom_results(sess), "scores.csv")
```

That CSV has everything: document IDs, concept names, scores, the model's
rationales, and highlight quotes.

For analysis you usually also want the **wide** version — one row per
document, one column per concept — to join onto your main dataset.
[scores_wide()] does the pivot and sanitizes concept names into valid
column names (continuing with the synthetic scores from above):

```{r save-wide}
wide <- scores_wide(score_df, "doc_id")
wide

# The mapping from concept names to column names:
attr(wide, "concept_names")
```

Joining the scores back onto your data is then a regular merge — with the
row-count checks that any merge deserves:

```{r save-join}
stopifnot(nrow(wide) == nrow(docs))
merged <- dplyr::left_join(docs, wide, by = "doc_id")
stopifnot(nrow(merged) == nrow(docs))
head(merged, 3)
```

Finally, two one-liners for whole-analysis persistence:

```{r save-all, eval = FALSE}
# Everything at once: scores (long + wide), concepts, evidence table,
# and the full session object
lloom_write(sess, "results/")

# Or just the session; chat objects survive and keep working after reload
saveRDS(sess, "results/session.rds")
sess <- readRDS("results/session.rds")   # later, in a fresh R session
```

## Further reading

- Lam, M., et al. (2024). *Concept Induction: Analyzing Unstructured Text
  with High-Level Concepts Using LLooM.* CHI 2024.
  <https://doi.org/10.1145/3613904.3642830>
- Upstream Python package: <https://github.com/michelle123lam/lloom>
